---
title: "data_01_create_bill_pay_df"
date: "2023-01-03"
---

# Overview

Before creating the Shiny app, I want to work through some of the code that will be used on the back end to merge together and update the data tables.

In this file, I'm preprocessing and merging the separate data tables needed to create the `bill_pay` data frame. I'm thinking this will be the table I use when I'm paying bills.

[Here is a link](https://drawsql.app/teams/brad-cannells-team/diagrams/budget-bill-pay-view) to the SQL diagram.

I find using Qmd files easier to work with while developing than R scripts. However, the code used by the Shiny app will be in an R script. 

# Load packages

```{r}
library(dplyr, warn.conflicts = FALSE)
library(readr)
```


# Load data tables

```{r message=FALSE}
entity          <- read_csv("../tables/entity.csv")
account         <- read_csv("../tables/account.csv")
interest_rate   <- read_csv("../tables/interest_rate.csv")
payment_method  <- read_csv("../tables/payment_method.csv")
period_payment  <- read_csv("../tables/period_payment.csv")
```


# Get current date

We will use this to filter table elements to display

```{r}
today <- Sys.Date()
```


# Data management

## Subset the entity table

Only keep the entities that I would potentially be paying in this table.

Also, remove any closed accounts.

## Merge the data frames for bill payment together

```{r}
pay_bills <- entity |> 
  left_join(account, by = c("pk_entity" = "fk_entity"), keep = TRUE) |> 
  left_join(interest_rate, by = c("pk_account" = "fk_account"), keep = TRUE) |> 
  left_join(payment_method, by = c("pk_account" = "fk_account"), keep = TRUE) |> 
  left_join(period_payment, by = c("pk_account" = "fk_account"), keep = TRUE)
```

Add an primary key column to the pay_bills data frame. 

```{r}
pay_bills <- pay_bills |> mutate(pk_pay_bills = row_number())
```

```{r}
dim(pay_bills) # 5 53
```

# Keep variables of interest

For the bill pay view, we don't want to display all 52 columns. Instead, we more or less want to display the handful of columns that are shown on my Google Sheet budget.

```{r}
pay_bills_subset <- pay_bills |> 
  select(
    pk_pay_bills, entity_name, entity_url, period_payment_balance, 
    period_payment_payment_expected,
    period_payment_payment_actual, period_payment_payment_diff_c, 
    period_payment_paid, period_payment_pending, period_payment_notes
  )
```

```{r}
dim(pay_bills_subset) # 5 10
```

# Add data

Now, let's add data to the pay_bills_subset table as if we were paying bills.

Later, we will need to add the new information back to the the larger pay_bills table somehow, and then into the individual, stored data tables too. 


```{r}
#| purl: false

# Convert to R script
# So that it can just be sourced by app.R. 
# Wrote these notes as comments inside the code chunk so that they don't show up in the R script. 

knitr::purl("data_01_create_bill_pay_df.qmd", "../R/data_01_create_bill_pay_df.R", documentation = 2)
```