---
title: "data_01_create_bill_pay_df"
date: "2023-01-03"
---

# Overview

In this file, I'm preprocessing and merging the separate data tables needed to create the `bill_pay` data frame.

I find using Qmd files easier to work with while developing than R scripts. However, the code used by the Shiny app will be in an R script. 

# Load packages

```{r}
library(dplyr)
library(readr)
```


# Load data tables
```{r}
payee_details <- read_csv("tables/payee_details.csv")
payees        <- read_csv("tables/payees.csv")
payments      <- read_csv("tables/payments.csv")
```


# Get current date

We will use this to filter table elements to display
```{r}
today <- Sys.Date()
```


# Data management

## Merge the data frames for bill payment together
```{r}
pay_bills <- payees |> 
  left_join(payee_details, by = c("pk_payees" = "fk_payees")) |> 
  left_join(payments, by = c("pk_payee_details" = "fk_payee_details"))
```

## Keep only the columns of interest

# Convert to R script

So that it can just be sourced by app.R. 

Not currently working.

```{r purl=FALSE}
knitr::purl("data_01_create_bill_pay_df,qmd", "../R/data_01_create_bill_pay_df.R", documentation = 2)
```