---
title: "Develop Bill Pay Logic"
format: html
date: "2024-03-22"
---

# Overview

Originally, I was doing all of this in bills_due.R. When the coding is complete, it will probably all have to go back into an R script for Shiny, but during the development phase, I think a Quarto file might be easier to work with.

I did a fair amount of work on the budget in Google Sheets. My thought was that I would just use the easiest platform to get the job done. It turns out that Google Sheets isn't "easy"; although, it still may turn out to be the easiest - we'll see. For now, however, I want to give Shiny another try.

One of the sheets that I started creating in Google Sheets was the "Bills Due" table. This table/sheet was supposed to be the view I would use for paying bills each pay period. The idea was for it to be similar to my traditional Google Sheets monthly budget, but dynamic.

In this file, we aren't yet creating a Shiny app. Instead, I just trying to create a static version of the bills due table that I want to interact with. Doing so requires joining data from a few different data tables and then filtering them based on the date and pay period.

Wiki about this topic: https://github.com/brad-cannell/budget_app/wiki/Pay-Bills

# Load packages

```{r}
library(dplyr, warn.conflicts = FALSE)
library(googlesheets4)
library(lubridate, warn.conflicts = FALSE)
```

# Import data

Save the sheet id for use below. You can pull it from the sheet's URL. For example, the sheet id for the budget app is: "1153fK76wz8cu4URFM7eidlyv8AEUoSKAAIt-q6uH0ZE" 

It is taken from the budget app's url, which is: "https://docs.google.com/spreadsheets/d/1153fK76wz8cu4URFM7eidlyv8AEUoSKAAIt-q6uH0ZE/edit#gid=1694905897"

```{r}
test_data_id <- "1UynDArMMcKkiJevSZtW4HJ74zMiqrLuKg0CpDcPtI2c"
test_accounts <- read_sheet(
  test_data_id,
  sheet = "Accounts"
)
```

# Selection parameters

These will be interactive in the Shiny app

## Get today's date
```{r}
today <- Sys.Date()
```

## Set year, month, and pay period

Originally, I was going to use today's year, but I should really be able to select any arbitrary year. Hopefully, I can set the default value to today's year somehow and then change it if needed. Again, these will be interactive in Shiny.

```{r}
selector_year <- 2024
selector_month <- "February"
selector_pay_period <- "Second"
```

# Step 1: Add a year filter

Ideally, I'd like to use this app across years. That means that there will be data from across multiple years in the data. So, I will need to be able to select a year I'm interested in (should default to the current year) and then view the bills that are due for the year and pay period selected.

So, if the account was opened on or before the selector date year and it was closed on or after the selector date year -- including accounts that  are still open (i.e., is.na(account_date_closed)).

```{r}
accounts_filtered_year <- test_accounts |>
  filter(year(account_date_opened) <= selector_year) |> 
  filter(is.na(account_date_closed) | year(account_date_closed) >= selector_year)
```

- Lake house should be filtered out. Opened after 2024.
- Old energy company should be filtered out. Closed in 2023.

```{r}
accounts_filtered_year
```

Step 2: Filter by selected month

Now, we have filtered our accounts list so that it only includes payment accounts that were open in the currently selected year. However, bills change  throughout the year. Now, we want to further filter the accounts list to only see accounts that are/were open in the currently selected month.

So, filter out accounts that were:  
- OPENDED AFTER the selected month (within the year) OR 
- CLOSED BEFORE the selected month (within the year). 

Said another way, KEEP accounts that were: 
- OPENDED BEFORE the selected month (within the year) OR 
- CLOSED AFTER the selected month (within the year).

## Get the month selected by the user as a number 1-12.

```{r}
selector_month_num <- match(selector_month, month.name)
```

In this example, the currently selected month is February. New credit card should be filtered out because it wasn't opened until March. Mattress should be filtered out because was closed in January.

```{r}
accounts_filtered_year |> 
  mutate(
    account_opened_month = month(account_date_opened),
    account_closed_month = month(account_date_closed)
  ) |> 
  select(account_nickname:account_date_opened, account_opened_month:account_closed_month) |> 
  filter(account_opened_month < selector_month_num | account_closed_month > selector_month_num)
```

## Issue

Trying to filter on the month number alone (as we did above) doesn't work. It filters out too many rows. For example, Therapy is filtered out because its open date month is 09, which is greater than 02. And if this account had been opened in the ninth month of 2024, that would be exactly what we want to  happen. However, Therapy was opened in 2021-09, so we don't want to filter it out.  

So, we need to filter by the month AND year the account was opened. If the selector year is 2024 and the selector month is February, then we want to keep all accounts that were open at any point during 2024-02. We will worry about filtering the pay period later. So,
- Account open date before 2024-02-01 AND 
- Account closed date on or after 2024-02-29

## Year-Month filters

```{r}
selector_month <- "February" # Adding again here for testing
```

Build selection date value from the selection parameters entered by the user.
Get the month selected by the user as a number 1-12. This will help us build a completed sector date (Y-M-D) below.
selector_month_num <- match(selector_month, month.name)

Merge together the year and month selected by the user as a date. We will use the first day of the month (i.e., "1") for now.

```{r}
selector_year_month_first_day <- paste(selector_year, selector_month_num, "1", sep = "-") |> as.Date()
```

Also determine what the last day of the selected month is.

```{r}
selector_year_month_last_day <- ceiling_date(selector_year_month_first_day, "month") - days(1)
```

## Filter by selected year and month

In this example, the currently selected month is February (2024). 
- New credit card should be filtered out because it wasn't opened until March.
- Mattress should be filtered out because was closed in January.

```{r}
accounts_filtered_year_month <- accounts_filtered_year |> 
  # Keep accounts opened before the last day of the selected year and month.
  # This should remove New credit card.
  filter(account_date_opened <= selector_year_month_last_day) |> 
  # Keep accounts that haven't been closed yet OR were closed after the first
  # day of the selected year and month.
  # This should remove Mattress.
  filter(is.na(account_date_closed) | (account_date_closed >= selector_year_month_first_day))

accounts_filtered_year_month
```

This works. 
If we change selector_month to January, Mattress should not be removed. I did a test and it works.

# Step 3: Filter by pay period

At this point, every account remaining was open at some point during the currently selected year and month. Now, we need to filter for the selected
pay period.

```{r}
selector_pay_period <- "First" # Adding again here for testing
```

```{r}
accounts_filtered_year_month |> 
  filter(account_pay_period_manual == selector_pay_period)
```

## Issue

The code above works just fine if we only have accounts that need to be paid in the first OR second pay period, MONTHLY. However, we have some accounts that need to be paid twice a month and other accounts that need to be paid annually. How do we handle them?

